\documentclass[11pt,c]{article}
\usepackage{amsmath}
\usepackage{amsfonts}
% \usepackage{mathrsfs}
% \usepackage[draft]{graphicx}
\usepackage{graphicx}

\usepackage[english]{babel}
\usepackage{times}
\usepackage{a4wide}
\usepackage{float}
\usepackage[numbers]{natbib}
% \usepackage{showlabels}
\usepackage{amssymb}
\usepackage{amsbsy}
\usepackage{pgf}
\usepackage{tikz}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
%\usepackage{tikz-cd}
\usepackage[all,cmtip]{xy}
\usepackage{caption}
\usepackage{subcaption}
% \captionsetup[figure]{labelformat=empty}% redefines the caption setup of the figures environment in the beamer class.

\input{my_preamble}
\newcommand{\red}[1]{\textcolor{red}{#1}}



% COLORS
\def\Reddd{\special{color rgb  1.   0.   0.}}
\def\Black{\special{color cmyk 0.   0.   0    1.}}
\let\B\Black
\let\Rd\Reddd




\catcode`@=12

             \topmargin 0.0in      %(TOP)  1"
             \oddsidemargin 0.0in  %(LEFT) 1"
             \textwidth 6.5in      %(RIGHT)1"
             \textheight 8.5in     %(BOT)  1"
             \headsep 0.0in
             \headheight 0.0in



\newcommand{\PT}{{\partial T}}
\newcommand{\sth}{{\textsuperscript{th}}~}

\let\tilde\widetilde

\setcounter{secnumdepth}{5}
\begin{document}

\begin{center}


{\huge {\bf
An Adaptive DPG Method for High Frequency Time-harmonic Wave Propagation Problems
\\}}
% \vspace*{1cm}

\vspace*{0.9cm}
{\Large
Socratis Petrides, Leszek F. Demkowicz

\vspace*{.5cm}

{\bf Institute for
Computational Engineering and Sciences\\
The University of Texas at Austin, Austin, TX 78712, USA\\}
}
\end{center}


\begin{abstract}

The Discontinuous Petrov-Galerkin (DPG) method for high frequency wave propagation problems is discussed. The DPG method offers uniform pre-asymptotic stability for any wavenumber. This allows for a fully automatic adaptive $hp$ algorithm, that can start from very coarse meshes. Moreover, DPG always delivers a Hermitian positive definite system, suggesting the use of the Conjugate Gradient algorithm for its solution. We present a new iterative solution scheme which benefits from these attractive properties of DPG. This novel solver is integrated within the adaptive procedure by constructing a two-grid-like preconditioner for the Conjugate Gradient method that  exploits information from previous meshes. The construction of such a preconditioner is discussed, and an example for a 2D acoustics problem is presented. Our results show that the proposed iterative algorithm converges at a rate independent of the mesh and the wavenumber.


\end{abstract}
\textbf{Keywords:} discontinuous test functions, finite elements, adaptivity, Petrov-Galerkin method, multigrid, preconditioner, iterative solvers
\baselineskip=16pt
\parskip= 4pt
%*************************************************page 1
\input{intro}
% \newpage
% \paragraph*{Acknowledgments.}
\newpage

\bibliography{bib}
\bibliographystyle{apalike}
\newpage
\appendix
\pagenumbering{arabic}
\renewcommand*{\thepage}{\thesection~\arabic{page}}

\end{document}

